### 2023-05-25 reports/paper/paper.qmd
get ready for SER

\fancyhead[R]{}
\fancyhead[L]{}
\renewcommand{\headrulewidth}{0pt}
\singlespacing

\thispagestyle{empty}

\setlength{\columnseprule}{0pt}

@@

**Background:** Non-Hodgkin lymphoma (NHL) incidence has increased substantially in the US since 1960 and was recently linked to workplace exposure to soluble metalworking fluids (MWF) in a standard survival analysis of the United Auto Workers-General Motors (UAW-GM) cohort. To further explore this association, we investigate the causal effects of hypothetical limits on soluble MWF exposure in relation to NHL risk in the same cohort of autoworkers.

**Methods:** We estimated counterfactual risk of NHL between 1985 and 2005 in the UAW-GM cohort of autoworkers at three Michigan plants (n = 34,734) under hypothetical limits on annual average daily exposure to soluble MWF. We applied the hazard-extended ICE parametric g-formula, an estimator capable of adjusting for time-varying confounding affected by past exposure.

**Results:** 231 NHL cases occurred over the 20-year follow-up period. Stronger hypothetical limits on soluble MWF exposure resulted in monotonic reductions in NHL risk. Capping annual average daily exposure at 0.5 mg/m^3^, the NIOSH recommended exposure limit, would have prevented 44 (95% CI: -6, 91) cases. Capping at 0.25 mg/m^3^ would have prevented 52 (95% CI: -5, 106) cases. Finally, capping at 0.05 mg/m^3^ would have prevented 71 (95% CI: 12, 129) cases.

**Conclusion:** Limits on soluble MWF exposure would have prevented cases of NHL in the UAW-GM cohort. Although MWF use is expected to decline as the automotive industry turns to electric vehicles, a boom in electric vehicle manufacturing could nonetheless lead to global increases in worker exposure to MWF. Strong limits on exposure to MWF may be important for preventing NHL incidence during a period of automotive industry expansion.

@@

In recent years, there has been a surge in optimism for growth in American manufacturing [@Carr_2022;@Armbrust_2021]. While it may be easy to imagine a factory floor crowded with workers and heavy machinery, an essential part of the industrial production of physical goods is easy to overlook: lubricants and coolants. In industries that require metalworking, from cutting wafers for semiconductor production to stamping metals for electric vehicle battery casings, metalworking fluids (MWF) are essential. During use in drawing, stamping, cutting, and grinding operations, MWF are sprayed and poured onto tools and parts to provide lubrication and cooling. Parts may even be submerged in MWF in precision manufacturing applications [@Arrabiyeh_2020].

Non-Hodgkin Lymphoma (NHL) is the seventh most common cancer type in the United States and has garnered substantial research attention because its incidence in the United States doubled between 1973 and 1994, before plateauing at around 19 per 100,000 persons per year [@seer_1994;@Ekstrom-Smedby_2006]. The strongest known risk factor of NHL is immunosuppression, both congenital and acquired [@Filipovich_1992;@Chiu_2015]. Modulation of immune function contributes to the pathogenesis of NHL directly by perturbing lymphocyte function or indirectly by reducing immunosurveillance against neoplasms. Immunodeficiency may also interact with infection with Epstein-Barr Virus, resulting in uncontrolled B-cell proliferation [@Oertel_2002]. Acquired immunodeficiency through infection with human immunodeficiency virus (HIV) can lead to genetic lesions and dysregulation while increasing vulnerability to infection by other oncogenic agents [@Grulich_1999;@Alexander_2007]. However, since immunodeficiency and infection with HIV are rare, they cannot fully explain the historic rise or present burden of NHL [@Shiels_2013]. Since the rise in NHL incidence coincided with a period of rapid and extensive chemicalization in industry, agriculture and warfare, there has been stronger interest in the environmental and occupational epidemiology of NHL [@Nelson_2005;@Romero_2021].

Pesticide exposure in agricultural occupational settings was a common target of NHL research in recent decades. A meta-analysis of 44 articles published between 1980 and 2014 found statistically significant associations between NHL and exposure to several classes of pesticides including carbamate, organophosporus, triazine, and organochlorine [@Schinasi_2014]. Occupations associated with NHL are not limited to the agricultural sector, however. Metal processors, health workers, salespeople, machinists, and electricians have all been identified as occupational groups with higher risk of NHL [@Fritschi_2005a; @Mester_2006; @Ekstrom-Smedby_2006].

Metalworking fluids (MWF), an important class of occupational exposures, are complex mixtures of oils and chemicals used to lubricate and cool metal tools and parts during metal cutting and grinding operations in manufacturing processes. During use, MWFs aerosolize and present a potential health hazard to exposed workers. There are three general types of MWF: straight, soluble, and synthetic. The type of MWF applied has changed over the decades and depends in part on the particular metalworking operation. Straight MWF, the oldest class of MWF, is comprised mostly of mineral oils, which were classified as carcinogenic by the International Agency for Research on Cancer (IARC) in 1973, citing studies of occupational exposures among workers in oil, textile, and metal industries [@IARC_1973]. The combination of health concerns, the high cost of oil in the 1970s, and the introduction of performance-enhancing water-soluble chemical additives led to the gradual replacement of straight MWF with water-based soluble and synthetic MWFs for many, but not all operations [@Childers_2005]. Although water-based MWF contain little to no mineral oil, they contain a richer cocktail of additives including chromates, cycloalkanes, phenols, organochlorines, nitrites, sulfonates, and triazines: classes of chemicals with known or plausible carcinogenicity [@IARC_1987;@IARC_2014;@IARC_2017].

In occupational studies, measures of time-varying exposure are essential to statistical adjustment for forms of time-varying confounding affected by past exposure such as the healthy worker survivor effect (HWSE). The HWSE is the dynamic selection process by which healthier individuals remain at work, where they accumulate more occupational exposure, and less healthy individuals leave work [@Arrighi_1994]. Standard analyses of occupational cohorts affected by the HWSE result in a downward bias on the exposure-outcome associations of interest, as less healthy workers typically accrue less exposure than healthier workers.

Here, we assess the effect of hypothetical MWF exposure limits on cumulative incidence of NHL from 1985 to 2005 in the United Auto Workers-General Motors (UAW-GM) occupational cohort study.
A recent study of cancer incidence in this cohort linked NHL with exposure to soluble MWF using a standard survival analysis using Cox proportional hazard regression [@Colbeth_2023]. Standard regression methods are vulnerable to bias due to the HWSE. We therefore further explore this association using an estimator capable of adjusting for time-varying confounding and censoring affected by past exposure: the hazard-extended iterative conditional expectation (ICE) parametric g-formula. Like other causal estimators in statistics, the hazard-extended ICE g-formula yields marginal effect estimates of hypothetical interventions, which may be more interpretable under a population health framework than conditional measures.

@@

We estimated NHL cumulative incidence from 1985 to 2005 under hypothetical limits on annual average daily exposure to soluble MWF by applying the hazard-extended ICE parametric g-formula estimator [@Wen_2020]. First, we estimated the expected number of NHL cases that we would observe if there were no censoring by competing risks. Then, we contrasted this counterfactual cumulative incidence to that under three hypothetical exposure limits:

(1) 0.5 mg/m^3^: the National Institute for Occupational Safety and Health (NIOSH) Recommended Exposure Limit (REL) for total particulate mass (PM) composed of MWFs [@niosh_1998]
(2) 0.25 mg/m^3^: Half the REL
(3) 0.05 mg/m^3^: One tenth of the REL.

Exposure limits were enforced for soluble MWF over workers' entire working lifetimes, thereby inducing a post-intervention exposure distribution from which exposures are drawn.

@@

We used data from the UAW-GM cohort, which includes all hourly workers at three automobile manufacturing plants in Michigan who had worked at least three years by 1985. Past papers provide detailed descriptions of the cohort [@Eisen_1992;@Eisen_2001]. The large size of the study population and rich time-varying, quantitative MWF exposure data provide an opportunity to study a relatively rare cancer and evaluate realistic interventions on MWF exposure in a longitudinal cohort setting. The present study population (N = 34,738) was restricted to the autoworkers who were at work in 1941 or not yet hired, missing no more than half of their employment history, and still alive at the start of follow-up. Autoworkers in the study population were followed for NHL incidence from January 1, 1985 until NHL diagnosis, death, January 1, 2005 (10 years after the end of employment record availability), or age 108 years (the oldest observed age at death), whichever came earlier.

@@

We identified incident cancers in the UAW-GM cohort that occurred between 1985 and 2005 by linkage to the Michigan Cancer Registry (MCR). Workers at Plants 1 and 2, located in the greater Detroit metropolitan area, were also linked to the Detroit Regional Registry of the Surveillance, Epidemiology, and End Results (SEER) Program. Cancer types were distinguished using site and histology codes conforming to the International classification of Diseases for Oncology, 3^rd^ edition (ICD-O-3). Non-Hodgkin lymphoma was defined by cancers with any of the following ICD-O-3 Histology codes: 9590-9597, 9670-9671, 9673, 9675, 9678-9680, 9684, 9687-9691, 9695, 9698-9702, 9705, 9708-9709, 9712, 9714-9719, 9724-9729, 9735, 9737-9738, 9811-9818, 9823, 9827, 9837. Details regarding cancer incidence follow-up are described elsewhere [@Colbeth_2023]. Vital status was ascertained from company records and by linkage to Social Security Administration, National Death Index, and state mortality files.

@@

Quantitative measurement of time-varying MWF exposure is a distinguishing strength of the UAW-GM cohort study relative to other occupational cohort studies. Exposure assessment was based on direct air sampling as well as company records. Company industrial hygienists collected several hundred personal and area samples for total particulate matter (mg/m^3^) composed of MWF over many decades. Research industrial hygienists collected additional air sampling data when the cohort study was launched in the mid 1980s. These additional data were combined with the historical data and company records to construct a job-exposure matrix of quantitative 8-hour time-weighted average daily exposure estimates to soluble, straight, and synthetic MWFs for each combination of job, department, and plant over time. Workers' time-weighted annual average daily exposure to each MWF type was determined by combining this job-exposure matrix with employment records, which recorded time-varying job type, department, and plant for each employee from hire to termination or 1994, whichever came sooner. For employment records that were at least half complete, gaps in the record were interpolated by carrying forward the last known job type. The exposure assessment is described in detail elsewhere [@Hallock_1994;@Woskie_1994;@Woskie_2003]. Previous analyses of NHL applied exposure lags of 1 to 20 years to account for disease latency; we lagged cumulative MWF exposures by 10 years and therefore ended follow-up on January 1, 2005 [@Smith_2007;@Karipidis_2007;@Zhang_2019]. In analyses, MWF exposure history was summarized as the cumulative sum of annual exposure intensities and coded as categorical variables with cut-points determined according the quantiles of cumulative exposure among cases. In analyses, we estimated the effects of interventions on soluble MWF while treating exposure to straight and synthetic MWFs as potential confounders.

@@

The hazard-extended ICE parametric g-formula may be thought of as a series of model-based standardization steps, which we implemented using logistic regression. We split the 20-year follow-up period into eight time periods; the first two periods spanned four years each, and the remaining six periods spanned two years each. The first two periods are longer in length to account for the smaller number of cases in those years. Post-intervention estimates of the discrete hazard of NHL given all exposures and covariates were combined iteratively from the end of follow-up to the start. In each iteration, predicted discrete hazards were standardized over post-intervention exposure and covariate histories before combining with discrete hazards from the previous iteration. This iterative process results in sequentially standardized estimates of NHL cumulative incidence over the entire follow-up period. Averaging over the baseline distribution of covariates yields the counterfactual cumulative incidence estimate of NHL if the intervention of interest were enforced for the entire study population.

Post-intervention exposure to soluble MWF and history of exposure to soluble MWF were summarized as cumulative exposure. We modeled discrete hazards by fitting a pooled logistic regression for NHL over at-risk person-periods given cumulative exposure to straight, soluble, and synthetic MWFs, employment status, cumulative time off, year of hire, sex (male/female), race (Black/white/unknown), and plant (Plant 1/Plant 2/Plant 3). Cumulative exposure to MWFs, employment status, and cumulative time off were lagged 10 years. All continuous variables were represented as categorical variables with cut points determined by the tertiles of nonzero values among NHL cases. An overview of the general steps of the estimation procedure are presented below.

0.  For the hypothetical exposure limit of interest, compute the cumulative exposure
		to soluble MWF that each worker would have accrued by the end of each follow-up
		period since hire.
1.  Fit a pooled logistic regression to the observed data for NHL on covariates and
    cumulative exposure to soluble MWF over all at-risk person-periods, excluding 
    those ending with a .
2.  Replace the cumulative exposure vector in the observed data with the vector
		of post-intervention cumulative exposure to soluble MWF. Using the model fitted
		in the  previous step, compute estimates of the post-intervention discrete hazard
		for each at-risk person-period including those that end with a censoring event.
3.  Set $k = K - 1$ where $K = 8$ is the number of follow-up periods:

	a) Among those who were event-free and uncensored through the $k$^th^ period,
		 fit a logistic regression on the predicted post-intervention discrete hazard
		 for the period spanning the $(k + 1)$ through $K$^th^ periods given observed
		 covariate and exposure history up through (and including) the $k$^th^ period.
	b) Replace the vector of cumulative exposure to soluble MWF in the observed data
		 with the post-intervention vector of cumulative exposure to soluble MWF. Using
		 the model fitted in the previous step, obtain predicted values for all those
		 who were at-risk in the $k$^th^ period.
	c) Compute the predicted post-intervention discrete hazard for the period spanning
		 the $k$ through $K$^th^ periods by multiplying the predicted values from the
		 previous step by 1 minus the discrete hazard estimate used to fit the model
		 in step 3.a and adding that same discrete hazard estimate to the product.
	d) If $k > 1$, set $k = k - 1$ and return to step 3.a.

4.  Compute the counterfactual risk by averaging the predicted post-intervention
		discrete hazards for the entire follow-up period for all units.

We estimated cumulative incidence under the observed distribution of soluble MWF exposure (natural course) and under the six interventions. We contrasted the cumulative incidence under intervention to that under the natural course by computing relative cumulative incidences. Confidence intervals were computed using the nonparametric bootstrap with 1000 Monte Carlo samples from the population at cumulative incidence at start of follow-up and centering on the estimate computed from observed data. All the necessary script used to reproduce the analyses are available on [GitHub](https://github.com/kvntchn/gm-nhl-ice).

@@

```{r}
load(here::here("resources", paste0("estimates.rdata")))
sapply(estimates, mean) 
estimates.bs <- readRDS(here::here("resources", paste0("estimates.bs.rds")))
```

@@

# Get lower and upper 2.5th %tiles
N <- 33134
rd <- rd[,.(
	rd = estimate[1],
	rd_lower = (2 * estimate[1] - quantile(value, 1 - 0.025)) * N,
	rd_upper = (2 * estimate[1] - quantile(value, 0.025)) * N
	# wald.lower = exp(log(estimate[1]) + qnorm(0.025) * sd(log(value))),
	# wald.upper = exp(log(estimate[1]) - qnorm(0.025) * sd(log(value)))
),
variable]
```

@@

# saveRDS(sol.tab, here::here('resources', 'sol.tab.rds'))

>>>>

@@

**Background:** Non-Hodgkin lymphoma (NHL) incidence has increased substantially in the US since 1960 and was recently linked to workplace exposure to soluble metalworking fluids (MWF) in a standard survival analysis of the United Auto Workers-General Motors (UAW-GM) cohort. To further explore this association, we estimate the effects of hypothetical limits on soluble MWF exposure in relation to NHL risk in the same cohort of autoworkers.

**Methods:** We estimated counterfactual risk of NHL between 1985 and 2015 in the UAW-GM cohort of autoworkers at three Michigan plants (n = 33,134) under hypothetical limits on annual average daily exposure to soluble MWF. We applied the hazard-extended ICE parametric g-formula, an estimator capable of adjusting for time-varying confounding affected by past exposure.

**Results:** 339 NHL cases occurred over the 30-year follow-up period. Stronger hypothetical limits on soluble MWF exposure resulted in monotonic reductions in NHL risk. Capping annual average daily exposure at 2 mg/m^3^ would have prevented 98 (95% CI: 45, 161) cases of NHL. Capping at 0.5 mg/m^3^, the NIOSH recommended exposure limit, would have prevented 112 (95% CI: 57, 181) cases. Finally, capping at 0.05 mg/m^3^ would have prevented 129 (95% CI: 66, 214) cases.

**Conclusion:** Limits on soluble MWF exposure would have prevented cases of NHL in the UAW-GM cohort. Although MWF use is expected to decline as the automotive industry turns to electric vehicles, a boom in electric vehicle manufacturing could nonetheless lead to global increases in worker exposure to MWF. Maintaining limits on exposure to MWF may protect workers against risk of NHL.

@@

<!-- In recent years, there has been a surge in optimism for growth in American manufacturing [@Carr_2022;@Armbrust_2021]. While it may be easy to imagine a factory floor crowded with workers and heavy machinery, an essential part of the industrial production of physical goods is easy to overlook: lubricants and coolants. In industries that require metalworking, from cutting wafers for semiconductor production to stamping metals for electric vehicle battery casings, metalworking fluids (MWF) are essential. During use in drawing, stamping, cutting, and grinding operations, MWF are sprayed and poured onto tools and parts to provide lubrication and cooling. Parts may even be submerged in MWF in precision manufacturing applications [@Arrabiyeh_2020]. -->

Non-Hodgkin Lymphoma (NHL) incidence in the United States doubled between 1973 and 1994, before plateauing at around 19 per 100,000 persons per year, making it the seventh most common cancer in the country [@seer_1994;@Ekstrom-Smedby_2006]. The strongest known risk factor of NHL is immunosuppression, both congenital and acquired [@Filipovich_1992;@Chiu_2015]. Modulation of immune function contributes to the pathogenesis of NHL directly by perturbing lymphocyte function or indirectly by reducing immunosurveillance against neoplasms. Immunodeficiency may also interact with infection with Epstein-Barr Virus, resulting in uncontrolled B-cell proliferation [@Oertel_2002]. Acquired immunodeficiency through infection with human immunodeficiency virus (HIV) can lead to genetic lesions and dysregulation while increasing vulnerability to infection by other oncogenic agents [@Grulich_1999;@Alexander_2007]. However, since immunodeficiency and infection with HIV are rare, they cannot fully explain the historic rise or present burden of NHL [@Shiels_2013]. The rise in NHL incidence coincided with a period of rapid and extensive chemicalization in industry, agriculture and warfare, suggesting that environmental and occupational exposures may have an important explanatory role in the epidemiology of NHL [@Nelson_2005;@Romero_2021].

Pesticide exposure in agricultural occupational settings was a common target of NHL research in recent decades. A meta-analysis of 44 articles published between 1980 and 2014 found statistically significant associations between NHL and exposure to several classes of pesticides including carbamate, organophosporus, triazine, and organochlorine [@Schinasi_2014]. Occupations associated with NHL are not limited to the agricultural sector, however. Metal processors, health workers, salespeople, machinists, and electricians have all been identified as occupational groups with higher risk of NHL [@Fritschi_2005a; @Mester_2006; @Ekstrom-Smedby_2006]. Workers in these occupational groups often come into contact with industrial chemicals such as gasoline, solvents, coolants, and lubricants such as metalworking fluids.

Metalworking fluids (MWF) are complex mixtures of oil, water, and chemical additives that optimize metal machining operations. During shaping, grinding, and cutting operations, MWF are misted, poured, or blasted at high pressure onto work surfaces to remove debris, cool metal, improve efficiency, and prevent deterioration of tools. Although MWFs provide essential engineering benefits to manufacturing processes, they may also present a potential health hazard to exposed workers, who may inhale or ingest aerosolized MWFs. There are three general types of MWF: straight, soluble, and synthetic. The type of MWF applied has changed over the decades in response to engineering, economic, and health considerations.

Straight MWF, the oldest class of MWF, is comprised mostly of mineral oils, which were classified as carcinogenic by the International Agency for Research on Cancer (IARC) in 1973, citing studies of occupational exposures among workers in oil, textile, and metal industries [@IARC_1973]. In the same decade, additional factors including the high cost of oil and the introduction of performance-enhancing water-soluble chemical additives spurred the gradual replacement of straight MWF with water-based soluble and synthetic MWFs for many, but not all metalworking operations [@Childers_2005]. Although water-based MWF contain little to no mineral oil, they contain a richer cocktail of additives including chromates, cycloalkanes, phenols, organochlorines, nitrites, sulfonates, and triazines: classes of chemicals with known or plausible carcinogenicity [@IARC_1987;@IARC_2014;@IARC_2017]. With policy makers and companies anticipating a boom in domestic manufacturing, occupational exposure to MWFs will remain an important concern for worker health in the US [@Armbrust_2021;@Carr_2022;@Keilman_2023].

Here, we assess the effect of hypothetical MWF exposure limits on cumulative incidence of NHL from 1985 to 2014 in the United Auto Workers-General Motors (UAW-GM) occupational cohort study. A recent study of cancer incidence in this cohort linked NHL with exposure to soluble MWF in a Cox proportional hazards regression [@Colbeth_2023]. Standard regression methods are vulnerable to downward bias due to the healthy worker survivor effect (HWSE), the process by which healthier individuals remain at work where they accumulate more occupational exposure. Therefore, we further explored this association using an estimator capable of adjusting for the HWSE and other sources of time-varying confounding and censoring affected by past exposure: the hazard-extended iterative conditional expectation (ICE) parametric g-formula [@Arrighi_1994;@Wen_2021]. Like other causal estimators in statistics, the hazard-extended ICE g-formula yields marginal effect estimates of hypothetical interventions, which may be more interpretable under a population health framework than conditional measures.

<!-- The healthy worker survivor effect (HWSE) is a key concern in occupational studies. The HWSE is the dynamic selection process by which healthier individuals remain at work, where they accumulate more occupational exposure, and less healthy individuals leave work [@Arrighi_1994]. Standard analyses of occupational cohorts affected by the HWSE result in a downward bias on the exposure-outcome associations of interest, as less healthy workers typically accrue less exposure than healthier workers. One analytic approach for addressing potential bias due to the HWSE and other time-varying confounders affected by past exposure is the application of longitudinal causal estimators such as the parametric g-formula [@Taubman_2009]. -->

@@

We estimated NHL cumulative incidence from 1985 to 2014 under hypothetical limits on annual average daily exposure to soluble MWF by applying the hazard-extended ICE parametric g-formula estimator [@Wen_2020]. First, we estimated the expected number of NHL cases that we would observe if there were no censoring by competing risks. Then, we contrasted this counterfactual cumulative incidence to that under interventions enforcing one of five hypothetical exposure limits in addition to no censoring, when positivity allows. The five limits we evaluated were (1) 2.0 mg/m^3^, (2) 1.0 mg/m^3^, (3) 0.5 mg/m^3^, (4) 0.25 mg/m^3^, and 
(5) 0.05 mg/m^3^. The National Institute for Occupational Safety and Health (NIOSH) Recommended Exposure Limit (REL) for total particulate mass (PM) composed of MWFs is 0.5 mg/m^3^ [@niosh_1998]. These interventions reduce exposures above the hypothetical limit, but allow exposures below the limit to vary according to their natural distribution. To prevent extrapolation, interventions considered reduce above-limit exposures to the highest limit-abiding exposure observed among other workers in the study population with the same confounder and exposure history at that time. If no such peers existed, the above-limit exposures among those workers would retain their natural distribution.

<!-- These interventions are known as threshold interventions that depend on the natural value of exposure. In general, threshold interventions set exposure to the hypothetical limit of interest when the natural value of exposure is beyond the threshold. Otherwise, these interventions leave exposure at the natural value. Such interventions are analogous to longitudinal stochastic interventions where the intervention distribution of exposure is the distribution of exposure induced by the threshold intervention rule [@Taubman_2009;@Young_2014;@Diaz_2021]. -->

@@

We used data from the UAW-GM cohort, which included all hourly workers at three automobile manufacturing plants in Michigan who had worked at least three years by 1985. Past papers provide detailed descriptions of the cohort [@Eisen_1992;@Eisen_2001]. The large size of the study population and rich time-varying, quantitative MWF exposure data provide an opportunity to study a relatively rare cancer and evaluate realistic interventions on MWF exposure in a longitudinal cohort setting. The present study population (N = 33,134) was restricted to the autoworkers who were at work in 1941 or not yet hired, missing no more than half of their employment history, and still alive at the start of follow-up. Autoworkers in the study population were followed for NHL incidence from January 1, 1985 until NHL diagnosis, death, December 31, 2014 (20 years after the end of employment record availability), or the oldest observed age at death (108 years), whichever came earlier.

@@

We identified incident cancers in the UAW-GM cohort that occurred between 1985 and 2014 by linkage to the Michigan Cancer Registry (MCR). Workers at Plants 1 and 2, located in the greater Detroit metropolitan area, were also linked to the Detroit Regional Registry of the Surveillance, Epidemiology, and End Results (SEER) Program. Cancer types were distinguished using site and histology codes conforming to the International classification of Diseases for Oncology, 3^rd^ edition (ICD-O-3). Non-Hodgkin lymphoma was defined by cancers with any of the following ICD-O-3 Histology codes: 9590-9597, 9670-9671, 9673, 9675, 9678-9680, 9684, 9687-9691, 9695, 9698-9702, 9705, 9708-9709, 9712, 9714-9719, 9724-9729, 9735, 9737-9738, 9811-9818, 9823, 9827, 9837. Details regarding cancer incidence follow-up are described elsewhere [@Colbeth_2023]. Vital status was ascertained from company records and by linkage to Social Security Administration, National Death Index, and state mortality files.

@@

Quantitative measurement of time-varying MWF exposure is a distinguishing strength of the UAW-GM cohort study relative to other occupational cohort studies. Exposure assessment was based on direct air sampling as well as company records. Company industrial hygienists collected several hundred personal and area samples for total particulate matter (mg/m^3^) composed of MWF over many decades. Research industrial hygienists collected additional air sampling data when the cohort study was launched in the mid 1980s. These additional data were combined with the historical data and company records to construct a job-exposure matrix of quantitative 8-hour time-weighted average daily exposure estimates to soluble, straight, and synthetic MWFs for each combination of job, department, and plant over time. 
Workers' time-weighted annual average daily exposure to each MWF type was determined by combining this job-exposure matrix with employment records, which recorded time-varying job type, department, and plant for each employee from hire to termination or 1994, whichever came sooner. For employment records that were at least half complete, gaps in the record were interpolated by carrying forward the last known job type. The exposure assessment is described in detail elsewhere [@Hallock_1994;@Woskie_1994;@Woskie_2003].

Previous analyses of NHL applied exposure lags of 1 to 20 years to account for disease latency; we lagged cumulative MWF exposures by 20 years and therefore ended follow-up on January 1, 2015 [@Smith_2007;@Karipidis_2007;@Zhang_2019]. In analyses, MWF exposure at start of follow-up was summarized as the cumulative sum of annual exposure intensities. Exposure was coded as categorical variables with cut-points at zero and the quintiles of exposure among cases. In analyses, we estimated the effects of interventions on soluble MWF while treating exposure to straight and synthetic MWFs as potential confounders.

@@

We split the 30-year follow-up period into ten time periods based on the deciles of the dates of diagnosis of NHL. The number of years per period range from two  to four years. The hazard-extended ICE parametric g-formula involves two stages. In the first, we estimate counterfactual discrete hazards over the person-periods. In the second, we pool those estimates to estimate counterfactual risk over the entire follow-up period. During pooling, we iteratively combine estimates of the counterfactual hazard starting with the last two periods, then the last three, then the last four, and so forth. In each iteration, we perform model-based standardization over exposure and covariate histories before combining the counterfactual discrete hazard with that pooled over subsequent periods. This iterative process results in a sequentially standardized estimate of NHL cumulative incidence over the entire follow-up period under the post-intervention distribution of exposure.

At the start of follow-up and in the first stage of estimation, we summarized history of exposure to soluble MWF as the cumulative sum of average daily exposure. In the first stage, we fitted a pooled logistic regression for NHL over at-risk person-periods given cumulative exposure to straight, soluble, and synthetic MWFs, employment status, cumulative time off, age, duration of employment, sex (male/female), race (Black/white/unknown), and plant (Plant 1/Plant 2/Plant 3). Exposure to MWFs, employment status, time off, and duration of employment were lagged 20 years. All continuous variables were represented as categorical variables with cut points determined by the quantiles of nonzero values among NHL cases. In the second stage, we conducted model-based standardization using logistic regression on the entire set of exposure and covariate histories. An overview of the general steps of the estimation procedure are presented below. Note that we refer to discrete hazards simply as hazards.

1.  Fit a pooled logistic regression of NHL on covariates
		and exposure over all at-risk person-periods.
2.  Predict the hazard for each person-period for each possible level of exposure
		using the model from step 1.
3.  Within strata formed by unique combinations of covariates, obtain the
    post-intervention distribution of exposure by applying the intervention rule
    to the observed exposure distribution.
4.  Within strata, estimate the counterfactual hazard for each 
		person-period by taking a weighted sum of the predicted hazards from step 2.
5.  Starting at the penultimate period of follow-up, estimate the pooled
    counterfactual hazard:

    a)  Regress the counterfactual hazard pooled over all subsequent periods on
        exposure and covariate history at the current period.
    b)  Predict the pooled hazard for each person at risk at the start of the
    		period and each possible level of exposure using the model from (a).
    		Retrieve the predicted hazards from step 2. For each person and level
    		of exposure, multiply the predicted pooled hazard by one minus the 
    		corresponding predicted hazard and add the predicted hazard to the
    		product.
    c)  Within strata formed by exposure and covariate history, obtain the
        post-intervention distribution of exposure by applying the 
        intervention rule to the observed exposure distribution.
    d)  Within strata, estimate the counterfactual hazard pooled over the present
    		and subsequent periods for each person by taking a weighted sum of the
    		predicted pooled hazards from (b).
    e)  If the present period is not first period, set the reference period to the
    		preceding period and return to (a).

6.  Compute the counterfactual risk by averaging the counterfactual hazard over
		the entire follow-up period across all persons.

To account for censoring, fit the models in step 1 and step 5a among those who were
not censored and obtain predicted hazards for all at-risk person-periods, including
those that were censored.

We estimated cumulative incidence under the observed distribution of soluble MWF exposure (natural course) and under the six interventions. We contrasted the cumulative incidence under intervention to that under the natural course by computing the number of cases averted and the cumulative incidence ratios. Confidence intervals were computed using the basic nonparametric bootstrap with 1000 Monte Carlo samples from the population at the start of follow-up. All the necessary script used to reproduce the analyses are available on [GitHub](https://github.com/kvntchn/gm-nhl-ice).

@@

```{r}
load(here::here("resources", paste0("estimates.rdata")))
estimates.bs <- readRDS(here::here("resources", paste0("estimates.bs.rds")))
```

@@

# Get lower and upper 2.5th %tiles
N <- 33134
rd <- rd[,.(
  rd = - estimate[1],
  rd_lower = - (2 * estimate[1] - quantile(value, 0.025)) * N,
  rd_upper = - (2 * estimate[1] - quantile(value, 1 - 0.025)) * N
  # wald.lower = exp(log(estimate[1]) + qnorm(0.025) * sd(log(value))),
  # wald.upper = exp(log(estimate[1]) - qnorm(0.025) * sd(log(value)))
),
variable]
```

@@

# save(sol.tab, mwf.flextable, file = here::here('resources', 'sol.tab.rdata'))
